Unlocking Sensitivity for Visibility-based Power Spectrum Estimators of the 21\,cm Reionization Signal

### 2017-07-19 Rotation paper/draft_working.tex
new draft

Observational studies of reionization, including Gunn-Peterson measurements of quasi-stellar objects \citep{Fan2006} and Cosmic Microwave Background temperature and anisotropy measurements (CMB; \citealt{Planck2016}), the Kinetic Sunyaev-Zeldo'vich effect \citep{kszpatchy} and Lyman alpha galaxy observations \citep{mcquinnLyA} have given us indications of the rough time-frame of reionization, but only limited constraints on the finer spatial and temporal structures. A surge of recent radio-astronomical experiments of reionization focus on measuring the ``spin-flip'' transition of neutral
hydrogen with a characteristic wavelength of 21\,cm \citep{Furlanetto2006181,PritchardLoeb}.
The 21\,cm brightness temperature is a direct tracer of neutral hydrogen through the epoch of reionization, and thus by measuring the three dimensional distribution of the 21\,cm signal we measure the full temporal and spatial variations of this event. 
However, before realizing full-scale 21\,cm tomography, many current radio interferometric efforts
aim to measure the spatial power spectrum of 21\,cm brightness temperature fluctuations.
Current-generation instruments include the Donald C. Backer Precision Array for Probing
the Epoch of Reionization (PAPER; \citealt{Ali2015,paper32}), Murchison
Widefield Array (MWA; \citealt{Bowman2013, Tingay2013}), Low Frequency Array (LOFAR; \citealt{LOFAR}). Next-generation instruments include the Hydrogen Epoch of Reionization
Array (HERA; e.g. \citealt{HERA,HERAconfiguration,HERABEAM1,HERADISH2}) under construction, 
and the Square Kilometer Array Low (SKA-low; e.g. \citealt{SKA1}) currently in planning stages.

@@

Eqs. \eqref{eq:nur} and \eqref{eq:thinshell} show that observational angular and spectral dimensions together correspond to the three dimensional volume of cosmological interest. Moreover, recall that the power spectrum in Eq. \eqref{eq:Pofk} is proportional to the product of Fourier transforms of the temperature field. Since Eq. \eqref{eq:Vis1} resembles a Fourier transform along the angular directions, it is natural to perform a further Fourier transform along the radial, or equivalently, frequency axis. The transform, called delay-transform \citep{delay-transform}, has gained popularity recently since it has been shown that foregrounds are isolated in delay space \footnote{We discuss foreground isolation in detail in Section \ref{sec:chromaticity}.}. We define the delay-transformed visibility as:
\small
\begin{equation}
\begin{aligned}V(\boldsymbol{b},\tau) & =\int d\nu V_{\nu}(\boldsymbol{b})\phi(\nu)\exp\left(2\pi i\nu\tau\right),\\
 & =\int d\Omega d\nu B(\hat{\boldsymbol{s}},\nu)T(\hat{\boldsymbol{s}},\nu)\exp\left[-2\pi i\nu\left(\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}-\tau\right)\right]. 
\end{aligned}
\label{eq:Vb1}
\end{equation}
\normalsize
Here the delay $\tau$ is the Fourier dual of $\nu$. Eq. \eqref{eq:Vb1} expresses the delay-transformed visibility as
an integral over observational coordinates $\hat{\boldsymbol{s}}$ and $\nu$. For notational simplicity here and in the rest of this paper, we have defined the quantity
\[
B(\hat{\boldsymbol{s}},\nu) \equiv \frac{2k_B}{\lambda^2}\phi(\nu)A(\hat{\boldsymbol{s}},\nu). 
\label{eq:B}
\]

@@

\begin{figure}[H]
\includegraphics[width=1.1\linewidth]{rephs}

@@

At the time offset of maximum redundancy, we expect two baselines to have to same fringe pattern in both frequency and phase. Due to the time delay, however, the beam centers would be slightly shifted with respect to each other. In the top panels of Fig. \ref{fig:beamfringe}, we show the real parts of peak-normalized beam-fringe patterns. The top left and middle panels show the beam-fringe
patterns for the baseline classes \{2,0\} and \{2,1\}, offset by 0.0325 sidereal days. More specifically, displayed are the real parts of $B(\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\hat{\boldsymbol{s}}\cdot\boldsymbol{b}\right]$ and $B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right]$, respectively. The top right panel show their conjugate product $B^{*}(\hat{\boldsymbol{s}},\nu)B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[i2\pi\frac{\nu}{c}\left(\hat{\boldsymbol{s}}\cdot\boldsymbol{b}-\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right)-i\psi_{\nu}\right]$, where we have chosen global phase $\psi_{\nu}$ such that the peak of the conjugate product is real. The product of the beam fringe patterns in top right shows that the fringes
 cancel out as we expect. Comparing with Eq. \eqref{eq:Thetanu}, we see that $\Theta_\nu$ is the sum over the top right plot. Thus we show that at the optimal time offset the fringe patterns of the two baselines cancel out, leading to a substantial weight $\Theta_\nu$. The frequency-summed weight $\Theta$ is constructed with rephasing as described in Section \ref{sec:rephs} so that the conjugate product (top right panel) for different frequencies align in phase.

Similarly, we show the correlation of the two baselines in terms of their $uv$ beam spread overlap. The bottom left and middle panels are (peak-normalized) instantaneous $uv$ coverage of two baselines while the bottom right panel is their product. We see from the bottom right panel that the cross-multiplied baselines recover power concentrated at $u\approx15$. The product of the $uv$ beams spread shown in the bottom right panel should not be confused with the Fourier transform of the beam-fringe conjugate product shown in the top right panel.

@@

We define the estimator sensitivity to be the inverse of the total power spectrum noise variance $\Sigma_P^2$:
\begin{equation}
\rho \propto 1/\Sigma_P^2 \propto \rho_0^2\sum_{\boldsymbol{bb'}}^N\widetilde{\Theta}^2_{\boldsymbol{bb'}},
\end{equation}
where, if $\sigma_S^2$ and $\sigma_N^2$ are the characteristic signal and noise levels of a single-baseline visibility, $\rho_0\equiv\sigma_S^2/\sigma_N^2$ is the signal to noise ratio.

@@

\item $\Delta t_{\boldsymbol{bb'}}$ vs. $\widetilde{\Theta}_{\boldsymbol{bb'}}$:
This relation  is familiar from Fig. \ref{fig:sensplot}. The points at $\Delta t=0$ are the equivalent baselines. Note HERA arrays have fewer data points with high $\Delta t_{\boldsymbol{bb'}}$, due to their comparatively narrow beam.

@@

\item $\widetilde{\Theta}_{\boldsymbol{bb'}}$ vs. $\bar{b}$:
The general trend to be seen here is that longer baselines tend to have lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$. This is due to the lower multiplicity of longer baselines. The trend is particularly obvious in the linear structure for the HERA arrays, which are the equivalent baselines. Since they all have the unity $\Theta_{\boldsymbol{bb'}}$ by definition, the linearly decreasing trend of $\widetilde{\Theta}_{\boldsymbol{bb'}}$ is a direct measure of the baseline multiplicity structures of the HERA array configurations.

@@

Clearly not all baseline pairs give significant contribution to sensitivity. Due to the large number of baseline pairs, it may not be computationally feasible to include all pairs. Having quantified the sensitivity from a given pair of baselines, we study the cumulative sensitivity of the array depending on which baseline pairs we include. Evidently we should prefer the pairs with larger $\widetilde{\Theta}_{\boldsymbol{bb'}}$. In Fig. \ref{fig:osens} we plot $\rho$ against the minimum $\widetilde{\Theta}_{\boldsymbol{bb'}}$. $\rho(\widetilde{\Theta}_{min})$ is the sensitivity of the array when baseline-pairs that have $\widetilde{\Theta}_{\boldsymbol{bb'}}>\widetilde{\Theta}_{min}$ are included. The dashed lines represent the values when only the equivalent baseline-pairs are used.  We see as expected that in all cases using the nearly equivalent baselines lead to more and more significant improvements with lower $\widetilde{\Theta}_{min}$, or in other words when more baseline pairs are used. The small HERA37, with no gap (like in HERA350) or short nearly equivalent baselines (like in PAPER 128), will not benefit much from the nearly equivalent baselines. The maximum benefits for other cases are expected to be around $20\%$ to $60\%$. PAPER-128 is designed with highly redundant nearly equivalent baselines, and thus  these baselines start contributing at higher $\widetilde{\Theta}_{min}$, but the gapped HERA configurations will benefit even more from nearly equivalent baselines at low $\widetilde{\Theta}_{min}$ due to the presence of more classes of such pairs. Note that here we normalized $\rho$ such that the contribution of the top equivalent baseline pair, such as are 1. This plot therefore does not compare the absolute sensitivity across the different arrays. The stepwise pattern is characteristic of a regular grid; as we step to lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$ large groups of baseline pair classes get included in ``batch''. 
%\begin{centered}
\begin{figure}[H]
\includegraphics[width=\linewidth]{osens}
\caption{Sensitivity of redundant arrays as a function of the minimum effective weight. Dashed lines represent when only equivalent baseline pairs are used, while solid lines indicate use of both the equivalent and nearly equivalent baselines are used. The vertical axis is normalized independently for each array such that the contribution of the top single equivalent pair class is unity, and thus does not indicate a comparison of absolute sensitivity across the different arrays shown.  }
\label{fig:osens}
\end{figure}
%\end{centered}

>>>>

Observational studies of reionization, including Gunn-Peterson measurements of quasi-stellar objects \citep{Fan2006} and Cosmic Microwave Background temperature and anisotropy measurements (CMB; \citealt{Planck2016}), the Kinetic Sunyaev-Zeldo'vich effect \citep{ZahnKSZ, GeorgeKSZ, kszpatchy} and Lyman alpha galaxy observations \citep{Rhoads2017, Rhoads2012, mcquinnLyA} have given us indications of the rough time-frame of reionization, but only limited constraints on the finer spatial and temporal structures. A surge of recent radio-astronomical experiments of reionization focus on measuring the ``spin-flip'' transition of neutral
hydrogen with a characteristic wavelength of 21\,cm \citep{Furlanetto2006181,PritchardLoeb}.
The 21\,cm brightness temperature is a direct tracer of neutral hydrogen through the epoch of reionization, and thus by measuring the three dimensional distribution of the 21\,cm signal we measure the full temporal and spatial variations of this event. 
However, before realizing full-scale 21\,cm tomography, many current radio interferometric efforts
aim to measure the spatial power spectrum of 21\,cm brightness temperature fluctuations.
Current-generation instruments include the Donald C. Backer Precision Array for Probing
the Epoch of Reionization (PAPER; \citealt{Ali2015,paper32}), Murchison
Widefield Array (MWA; \citealt{Bowman2013, Tingay2013}), Low Frequency Array (LOFAR; \citealt{LOFAR}). Next-generation instruments include the Hydrogen Epoch of Reionization
Array (HERA; e.g. \citealt{HERA,HERAconfiguration,HERABEAM1,HERADISH2}) under construction, 
and the Square Kilometer Array Low (SKA-low; e.g. \citealt{skalow2015} and references therein) currently in planning stages.

@@

Eqs. \eqref{eq:nur} and \eqref{eq:thinshell} show that observational angular and spectral dimensions together correspond to the three dimensional volume of cosmological interest. Moreover, recall that the power spectrum in Eq. \eqref{eq:Pofk} is proportional to the product of Fourier transforms of the temperature field. Since Eq. \eqref{eq:Vis1} resembles a Fourier transform along the angular directions, it is natural to perform a further Fourier transform along the radial, or equivalently, frequency axis. The transform, called delay-transform \citep{delay-transform}, has gained popularity recently since it has been shown that foregrounds are isolated in delay space \footnote{We discuss foreground isolation in detail in Section \ref{sec:chromaticity}.}. We define the delay-transformed visibility as:
\small
\begin{equation}
\begin{aligned}V(\boldsymbol{b},\tau) & =\int d\nu V_{\nu}(\boldsymbol{b})\phi(\nu)\exp\left(2\pi i\nu\tau\right),\\
 & =\int d\Omega d\nu B(\hat{\boldsymbol{s}},\nu)T(\hat{\boldsymbol{s}},\nu)\exp\left[-2\pi i\nu\left(\frac{\boldsymbol{b}\cdot\hat{\boldsymbol{s}}}{c}-\tau\right)\right]. 
\end{aligned}
\label{eq:Vb1}
\end{equation}
\normalsize
Here the delay $\tau$ is the Fourier dual of $\nu$. Eq. \eqref{eq:Vb1} expresses the delay-transformed visibility as
an integral over observational coordinates $\hat{\boldsymbol{s}}$ and $\nu$. For notational simplicity here and in the rest of this paper, we have defined the quantity

\[
B(\hat{\boldsymbol{s}},\nu) \equiv \frac{2k_B}{\lambda^2}\phi(\nu)A(\hat{\boldsymbol{s}},\nu). 
\label{eq:B}
\]

@@

\begin{figure}[H]
\includegraphics[width=1.1\linewidth]{rephs_cm}

@@

At the time offset of maximum redundancy, we expect two baselines to have to same fringe pattern in both frequency and phase. Due to the time delay, however, the beam centers would be slightly shifted with respect to each other. In the top panels of Fig. \ref{fig:beamfringe}, we show the real parts of peak-normalized beam-fringe patterns. The top left and middle panels show the beam-fringe
patterns for the baseline classes \{2,0\} and \{2,1\}, offset by 0.0325 sidereal days. More specifically, displayed are the real parts of $g_b\equiv B(\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\hat{\boldsymbol{s}}\cdot\boldsymbol{b}\right]$ and $g_{b'}\equiv B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[-i2\pi\frac{\nu}{c}\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right]$, respectively. The top right panel show their conjugate product $g_{b}^{*}g_{b'} = B^{*}(\hat{\boldsymbol{s}},\nu)B(\boldsymbol{\Gamma}\hat{\boldsymbol{s}},\nu)\exp\left[i2\pi\frac{\nu}{c}\left(\hat{\boldsymbol{s}}\cdot\boldsymbol{b}-\boldsymbol{\Gamma}\hat{\boldsymbol{s}}\cdot\boldsymbol{b'}\right)-i\psi_{\nu}\right]$, where we have chosen global phase $\psi_{\nu}$ such that the peak of the conjugate product is real. The product of the beam fringe patterns in top right shows that the fringes
 cancel out as we expect. Comparing with Eq. \eqref{eq:Thetanu}, we see that $\Theta_\nu=\int d\Omega g_{b}^{*}g_{b'}$ is the sum over the image shown in the top right panel. Thus we show that at the optimal time offset the fringe patterns of the two baselines cancel out, leading to a substantial weight $\Theta_\nu$. The frequency-summed weight $\Theta$ is constructed with rephasing as described in Section \ref{sec:rephs} so that the conjugate product (top right panel) for different frequencies align in phase.

Similarly, we show the correlation of the two baselines in terms of their $uv$ beam spread overlap. The bottom left and middle panels are (peak-normalized) instantaneous $uv$ coverage of two baselines ($\widetilde{g}_{b}$ and $\widetilde{g}_{b'}$) while the bottom right panel is their product. We see from the bottom right panel that the cross-multiplied baselines recover power concentrated at $u\approx15$. The product of the $uv$ beams spread shown in the bottom right panel should not be confused with the Fourier transform of the beam-fringe conjugate product shown in the top right panel.

@@

We define the estimator sensitivity to be the square root of the inverse of the total power spectrum noise variance $\Sigma_P^2$:
\begin{equation}
\rho \propto 1/\Sigma_P \propto \rho_0\sqrt{\sum_{\boldsymbol{bb'}}^N\widetilde{\Theta}^2_{\boldsymbol{bb'}}},
\end{equation}
where, if $\sigma_S^2$ and $\sigma_N^2$ are the characteristic signal and noise levels of a single-baseline visibility, $\rho_0\equiv\sigma_S^2/\sigma_N^2$ is the signal to noise ratio.

@@

\item $\Delta t_{\boldsymbol{bb'}}$ vs. $\widetilde{\Theta}_{\boldsymbol{bb'}}$:
This relation  is familiar from Fig. \ref{fig:sensplot}. The points at $\Delta t=0$ are the equivalent baselines. Note HERA arrays have fewer data points with high $\Delta t_{\boldsymbol{bb'}}$, due to their comparatively narrow beam. Also note that all near-equivalent pairs for HERA arrays appear at $\widetilde{\Theta}_{\boldsymbol{bb'}}\lesssim 0.4$.

@@

\item $\widetilde{\Theta}_{\boldsymbol{bb'}}$ vs. $\bar{b}$:
In this plot the HERA arrays each show two superimposed structures.  The equivalent baselines, all having the same individual weight $\Theta_{\boldsymbol{bb'}}$, appear as elongated structures towards high $\widetilde{\Theta}_{\boldsymbol{bb'}}$. The narrowness of these structures is an indication of the approximately linear and one-to-one correspondence between baseline length and multiplicity. The wide structures spanning the whole range of $\bar{b}$ are the near-equivalent baseline pairs. We see again for HERA arrays all near-equivalent pairs appear at $\widetilde{\Theta}_{\boldsymbol{bb'}}\lesssim 0.4$. For PAPER-128, the existence of short near-equivalent pairs with high multiplicity partially fills the high $\widetilde{\Theta}_{\boldsymbol{bb'}}$ region. 
Note the general trend that longer baselines tend to have lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$. This is due to the lower multiplicity of longer baselines. The trend is particularly obvious in the linear structure for the HERA arrays, which are the equivalent baselines. Since they all have the unity $\Theta_{\boldsymbol{bb'}}$ by definition, the linearly decreasing trend of $\widetilde{\Theta}_{\boldsymbol{bb'}}$ is a direct measure of the baseline multiplicity structures of the HERA array configurations.

@@

Clearly not all baseline pairs give significant contribution to sensitivity. Due to the large number of baseline pairs, it may not be computationally feasible to include all pairs. Having quantified the sensitivity from a given pair of baselines, we study the cumulative sensitivity of the array depending on which baseline pairs we include. Evidently we should prefer the pairs with larger $\widetilde{\Theta}_{\boldsymbol{bb'}}$. In Fig. \ref{fig:osens} we plot $\rho$ against the minimum $\widetilde{\Theta}_{\boldsymbol{bb'}}$. $\rho(\widetilde{\Theta}_{min})$ is the sensitivity of the array when baseline-pairs that have $\widetilde{\Theta}_{\boldsymbol{bb'}}>\widetilde{\Theta}_{min}$ are included. The dashed lines represent the values when only the equivalent baseline-pairs are used.  We see as expected that in all cases using the nearly equivalent baselines lead to more and more significant improvements with lower $\widetilde{\Theta}_{min}$, or in other words when more baseline pairs are used. The small HERA37, with no gap (like in HERA350) or short nearly equivalent baselines (like in PAPER 128), will not benefit much from the nearly equivalent baselines. Near-equivalent baselines make an increasingly significant contribution to the total sensitivity as the number of array elements are increased. PAPER-128 is designed with highly redundant nearly equivalent baselines, and thus  these baselines start contributing at higher $\widetilde{\Theta}_{min}$, but the gapped HERA configurations will benefit even more from nearly equivalent baselines at low $\widetilde{\Theta}_{min}$ due to the presence of more classes of such pairs. Note that here we normalized sensitivity for each array as the fraction of the total sensitivity when all baseline pairs are used. The plot therefore does not compare absolute sensitivity across different arrays. The stepwise pattern seen in almost all cases are characteristic of a regular grid; as we step to lower $\widetilde{\Theta}_{\boldsymbol{bb'}}$ large groups of baseline pair classes get included in ``batch''. Thus the ``gapless'' HERA37, with more regularity in its antenna configurations than the larger counterparts, exhibit more step patterns in fractional sensitivity. 
%\begin{centered}
\begin{figure}[H]
\includegraphics[width=\linewidth]{osens}
\caption{Fractional sensitivity of redundant arrays as a function of the minimum effective weight. Dashed lines represent when only equivalent baseline pairs are used, while solid lines indicate use of both the equivalent and nearly equivalent baselines are used. The vertical axis is normalized independently for each array to the total sensitivity when all baseline pairs, equivalent and near-equivalent, are used.}
\label{fig:osens}
\end{figure}
%\end{centered}
